---
title: "an-analysts-guide-to-adls-gen-2-R"
author: [elizabethotoole](https://github.com/elizabethotoole)
format: html
editor: visual
---

# An Analyst's Guide to Azure Data Lake Storage Gen 2 using R

Author: [elizabethotoole](https://github.com/elizabethotoole)

Adapted from: <https://github.com/edmundhaacke/nhs-r-reporting/blob/main/examples/rstudio_datalake_connection_example.R>

This beginner-friendly notebook serves as a step-by-step guide for connecting to Azure Data Lake Storage (ADLS) Gen 2 using R and a shared access signature (SAS).

Join me as I walk you through the essentials, so you can focus on analysing data, not battling with storage setups. 🚀

Example use case: To connect to lake data to undertake analysis and create a product/report as an HTML output.

## 1. A Quick Overview of Fundamental Concepts

#### What is ADLS Gen 2?

Azure Data Lake Storage Gen 2 (ADLS Gen 2) is a cloud storage service from Microsoft Azure designed to store large amounts of data.

#### How is it structured?

-   **Storage account** - The highest level of storage and is where all your data is stored and accessed.

    | e.g. "[https://youraccount.blob.core.windows.net"](https://youraccount.blob.core.windows.net%22)

    The youraccount part of the URL should be replaced with your actual Azure Storage account name. The account_url will be used to authenticate and connect to your specific Azure Storage account.

-   **Containers**- Inside a storage account, data is organised into containers. Think of containers as "buckets" that hold your data.

    |  e.g. "your-container-name"

    Replace "your-container-name" with the actual name of the container where your files are stored.

-   **Folders and Subfolders**: ADLS Gen 2 allows you to organise data using folders, this differs to traditional storage which is typically stored in flat containers.

    | e.g. "your-container-name/raw_data/"

    You can have folders like raw_data/, processed_data/, or analytics/, and within these, additional subfolders like logs/, reports/, etc.

-   **Blobs**: These are the files themselves which are stored inside the containers, folders, or subfolders.

    | e.g. "raw_data/logs/jan_log.csv"

    Blobs (aka files) can be in various formats like CSV, Parquet, JSON, and even images and videos.

#### What is a Shared Access Signature (SAS)?

A SAS token provides a way to limit access to your data by granting specific permissions (e.g., read-only) for a defined time frame and is a way to share access to specific files without exposing sensitive credentials.

SAS tokens are created by those with administrative privileges to the Azure Storage account and are stored in a key vault. Analysts typically don't set up SAS tokens themselves but can request a SAS token to access data from the administrator accordingly.

## 2. Downloading data from ALDS Gen 2 ⬇️

::: callout-caution
Note: Before loading the data, it's crucial to consider the size of your dataset as this will influence how we load the data into a dataframe for analysis.

This guide is aimed at analysts seeking to connect to ADLS gen 2 storage from an IDE like RStudio in order to work with datasets that can fit into memory.

Working with larger datasets is out of scope for this guide, I have listed some useful information and resources below.

As an analyst, you are likely to access larger datasets through dedicated environments such as Databricks where your fellow data engineers have done a lot of the heavy lifting with spark configurations.

Take a look at ["The R Developer's Guide to Databricks"](https://www.databricks.com/sites/default/files/2025-02/developers-guide-R.pdf) for more information.
:::

### Step 1: Import the necessary libraries

```{r load libraries, include = FALSE}

# Import the necessary libraries
# ---------------------------------------------------------

if (!require("tidyverse")) install.packages("tidyverse")
if (!require("dplyr")) install.packages("dplyr")
if (!require("AzureStor")) install.packages("AzureStor")
if (!require("AzureKeyVault")) install.packages("AzureKeyVault")
if (!require("arrow")) install.packages("arrow")
if (!require("data.table")) install.packages("data.table")

```

### Step 2: Connect to Azure Blob Storage

```{r setup lake connection details, include = FALSE}

# Replace with address of your Azure Key Vault where the SAS tokens are contained
  key_vault_url <- "<insert_key_vault_address>"
  
# Replace with the sas token name set up by the org.
  secret_name <- "<insert_token_name>"
  
# Replace youraccount with the name of your Azure Blob Storage account.
  account_url <- "https://youraccount.blob.core.windows.net"
  
# Replace your-container-name with the name of your container in Azure Blob Storage.
  storage_container_name <- "your-container-name" 

# Now setup connection settings
  vault <- key_vault(key_vault_url)
  
  sas_token <- vault$secrets$get(secret_name)
   
  storage_account <- storage_endpoint(account_url, sas = sas_token$value)
  
  container <- storage_container(storage_account, storage_container_name)

```

### Step 3: Load Data from the Blob Storage

For datasets that can fit into memory, you can access ADLS gen 2 storage with the following code. Examples provided are for CSV and parquet file types but could be expanded to support other file formats.

#### Read in a CSV

```{r read csv files, include = FALSE}

# Read in a single CSV
  filename <- "PATH/TO/.csv"
  rawdata <- storage_download(container, filename, dest = NULL)
  df <- read_csv_arrow(rawdata)
  view(df)
  
# Multiple files (csv)
## first list the files within the lake destination specified
  filenames <- list_adls_files(
    container,
    dir = "/<PATH>/",
    info = c("name"),
    recursive = TRUE
    ) # true gets all files within the folder path

## Iterate through each file in files names and read into df
   df <- rbindlist(
     lapply(
       filenames, function(i){
         rawdata <- storage_download(container,i,dest = NULL)
         read_csv_arrow(rawdata)
         }
       )
     )

```

#### Read in parquet files

```{r read parquet files, include = FALSE}

# Read in a single parquet
  filename <- "PATH/TO/.parquet"
  rawdata <- storage_download(container, filename, dest = NULL)
  df <- read_parquet(rawdata)

# Multiple files (csv)
## first list the files within the lake destination specified
  filenames <- list_adls_files(
    container,
    dir = "/<PATH>/", # folder directory
    info = c("name"), 
    recursive = TRUE
    ) # true gets all file names within the folder path, including subfolders

## iterate through each file in files names and read into df  
   df <- rbindlist(
     lapply(
       filenames, function(i){
         rawdata <- storage_download(container,i,dest = NULL)
         read_parquet(rawdata)
         }
       )
     )
```

### Step 4: Now view your data

```{r view data, include = FALSE}
view(df)
```

## Uploading Data to Azure Blob Storage ⬆️

The below code allows us to upload a file from our local machine into Azure Blob Storage which we can then subsequently incorporate into analytical projects.

```{r upload data to ADLS gen 2, include = FALSE}

# locate the file you want to upload include file type
  local_file_path = "Documents/file_1.csv" 

# define where you want to upload your file include file type
  blob_name = "uploads/file1.csv"  # Path in the container

# Now upload the file
  upload_adls_file(
    filesystem = container, # defined earlier in step 2.
    src = local_file_path,
    dest = blob_name, 
    blocksize = 2^24,  # Chunk size for file upload (optional, default is 16MB)
    put_md5 = FALSE,   # Calculate and store MD5 hash (optional)
    use_azcopy = FALSE # Use AzCopy for faster upload (optional)
  )
    
```

## Further Reading 📚

-   [Shared Access Signatures (SAS Tokens)](https://learn.microsoft.com/en-us/azure/ai-services/translator/document-translation/how-to-guides/create-sas-tokens?tabs=Containers)
-   [R: Operations on an Azure Data Lake Storage Gen2 filesystem](https://search.r-project.org/CRAN/refmans/AzureStor/html/adls.html)
